---
title: "ATLAS JSON Cohort Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ATLAS JSON Cohort Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

CohortIntelligence accepts cohort definitions as ATLAS JSON files. The package
uses `CirceR` to compile the JSON into a single read-only SQL CTE query —
no write permissions, no temp tables, no pre-existing cohort table required.

## 1. Diagnose before launching

```{r diagnose}
library(CohortIntelligence)

connection_details <- DatabaseConnector::createConnectionDetails(
  dbms     = "spark",
  server   = "your-host/default",
  user     = "token",
  password = "your-pat",
  port     = 443,
  pathToDriver = "/path/to/jdbc/"
)
connection <- DatabaseConnector::connect(connection_details)

connector <- create_cohort_connector(
  connection   = connection,
  cdm_schema   = "your_cdm_schema",
  vocab_schema = "your_vocab_schema"
)

# Check concept coverage and candidate patient count (read-only)
json_path <- system.file("template", "T2DM.json",
                          package = "CohortIntelligence")
check_cohort_json(connector, json_path)
# Output:
#   JSON OK. Concept IDs: 443732, 201826, 4193704, 1503297
#   Vocabulary: 4 / 4 concept(s) found in CDM.
#   546 patients with >= 1 qualifying condition (pre-rules).
#   If fetch_cohort_from_json() returns 0, run with verbose = TRUE.
```

## 2. Understand the T2DM template criteria

The bundled `T2DM.json` requires ALL of:

1. T2DM condition code (SNOMED 443732 / 201826 / 4193704 + all descendants)
2. Age >= 18 at index date
3. **365 days of continuous prior observation** before the index date
4. **Metformin prescription within 0-365 days AFTER** the index date

The 365-day prior observation window is the most common reason the cohort
returns 0 patients in specialty-care CDMs.

## 3. Debug a 0-patient result

```{r debug}
# Inspect the generated SQL
fetch_cohort_from_json(connector, json_path, verbose = TRUE)
# Prints the full CTE SELECT to the console

# The CirceR SQL structure:
#   WITH vwXXXCodesets AS (...)       -- concept expansion
#   , vwXXXqualified_events AS (...)  -- primary events (age + obs window)
#   , vwXXXInclusion_0 AS (...)       -- metformin inclusion rule
#   ...
#   , vwXXXfinal_cohort AS (...)      -- ERA-collapsed final cohort
#   SELECT ... FROM vwXXXfinal_cohort
```

## 4. Launch with the JSON

```{r launch}
launch_cohort_intelligence(
  connection   = connection,
  cdm_schema   = "your_cdm_schema",
  vocab_schema = "your_vocab_schema",
  json_path    = json_path
)

# Cohort loads automatically -- no button click needed.
# Disconnect when done:
DatabaseConnector::disconnect(connection)
```

## 5. Use your own ATLAS export

Export any cohort from ATLAS: **Export > JSON**.

```{r custom}
launch_cohort_intelligence(
  connection   = connection,
  cdm_schema   = "schema",
  vocab_schema = "schema",
  json_path    = "path/to/my_cohort.json"
)
```

Any valid ATLAS JSON for OMOP CDM >= 5.0 works. The CTE approach handles
concept ancestor expansion (`includeDescendants = TRUE`) correctly via the
`#Codesets` setup section.

## 6. Supported databases

| DBMS | `dbms` value | Notes |
|---|---|---|
| Databricks / Spark | `"spark"` | Tested |
| PostgreSQL | `"postgresql"` | Standard |
| SQL Server | `"sql server"` | Standard |
| Amazon Redshift | `"redshift"` | Standard |
| Google BigQuery | `"bigquery"` | Standard |
| Snowflake | `"snowflake"` | Standard |

JDBC drivers: `DatabaseConnector::downloadJdbcDrivers("spark", pathToDriver="/")`
