---
title: "Exporting a Clinician Review Packet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting a Clinician Review Packet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

The clinician review packet is a self-contained HTML report designed for
multidisciplinary meetings between data scientists, clinicians, and
epidemiologists. It contains:

1. Disclaimer (hypothesis-generating, not clinical validation)
2. Cohort overview (n patients, demographics summary, data density)
3. Cluster profile summary (one paragraph per cluster, cautious labels)
4. Review sets summary (patient counts per group)
5. Temporal flags summary (flag type × severity × count)
6. Candidate hypotheses (top 10, sortable by effect size)
7. Recommended next steps and limitations

No rmarkdown, pandoc, or LaTeX required — the report is built with
`htmltools` only.

---

## 1. Via the dashboard (recommended)

In the **Hypothesis & Report** tab, click **Configure & Export**, choose
which sections to include and how many patients, then click **Export HTML**.

---

## 2. Programmatically

```{r export}
library(CohortIntelligence)

# Assuming you have already run the pipeline:
# connector, cohort_members, domain_data, person_data,
# feature_matrix, ml_results, rank_df, quilt_base available

temporal_flags <- detect_temporal_flags(
  cohort_members = cohort_members,
  domain_data    = domain_data
)

review_sets <- build_review_sets(
  rank_df         = rank_df,
  domain_activity = domain_activity,
  feature_matrix  = feature_matrix,
  ml_results      = ml_results,
  cohort_members  = cohort_members,
  temporal_flags  = temporal_flags
)

hypotheses <- generate_hypotheses(
  feature_matrix  = feature_matrix,
  ml_results      = ml_results,
  min_effect_size = 0.1,
  max_hypotheses  = 20L
)

export_clinician_review_packet(
  results = list(
    cohort_members   = cohort_members,
    rank_df          = rank_df,
    domain_data      = domain_data,
    person_data      = person_data,
    ml_results       = ml_results,
    hypotheses       = hypotheses,
    quilt_base       = quilt_base,
    temporal_flags   = temporal_flags,
    review_sets      = review_sets,
    cluster_profiles = NULL      # computed internally if NULL
  ),
  path        = "cohort_review_2024.html",
  cohort_name = "T2DM Metformin Cohort",
  n_patients  = 10L,
  include     = list(
    typical        = TRUE,
    outliers       = TRUE,
    sparse         = TRUE,
    temporal_flags = TRUE,
    hypotheses     = TRUE
  )
)
```

---

## 3. Report tone and disclaimers

Every report includes the following disclaimer (not removable):

> *"This report is hypothesis-generating and based on structured OMOP CDM
> evidence only. It does NOT replace formal chart review, clinical adjudication,
> or causal inference analysis. All findings are preliminary and require
> clinical validation before use in research or patient care."*

When presenting the report:

- Open with the **Limitations** section to set expectations
- Use cluster labels as conversation starters, not conclusions
- For each temporal flag, ask: *"Could this reflect a data capture issue
  rather than a clinical event?"*
- For each hypothesis, ask: *"Does this difference make clinical sense?
  Would you expect it from first principles?"*

---

## 4. CSV side-exports

```{r csv}
# Export tabular data alongside the HTML report
export_cohort_results(
  results = list(
    rank    = rank_df,
    hypotheses = hypotheses
  ),
  path     = "review_exports/",
  formats  = c("csv","rds"),
  overwrite = TRUE
)
# Writes:
#   patient_rankings.csv
#   hypotheses.csv
#   (+ .rds versions)
```

---

## 5. Building the packet data structure separately

```{r packet}
# Build the packet in memory first (useful for inspection)
packet <- build_clinician_review_packet(
  cohort_summary    = build_cohort_summary(cohort_members, person_data),
  cluster_profiles  = build_cluster_profiles(rank_df, domain_data,
                                              cohort_members, person_data),
  review_sets       = review_sets,
  temporal_flags    = temporal_flags,
  hypotheses        = hypotheses,
  selected_patients = rank_df$subject_id[rank_df$priority_tier == "high"][1:5],
  patient_timelines = list()
)

# Inspect
str(packet, max.level = 1)

# Render
export_clinician_review_packet(
  results     = packet,
  path        = "my_report.html",
  cohort_name = "My Cohort"
)
```
