---
title: "Interpreting Anomaly Scores, Temporal Flags, and Hypotheses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting Anomaly Scores, Temporal Flags, and Hypotheses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Core principle

> All outputs from CohortIntelligence are **hypothesis-generating structured
> observations** from OMOP CDM records. They are not clinical diagnoses,
> causal claims, or validated findings.

This vignette explains how to interpret the three most complex outputs:
anomaly scores, temporal rule flags, and cluster-pair hypotheses.

---

## 1. Anomaly scores

### What they measure

Anomaly scores (0-1) come from an **isolation forest** — a tree-based algorithm
that isolates patients whose feature combinations are rare relative to the
cohort. A high score means the patient's pattern of conditions, drugs,
procedures, and visits across time windows is difficult to describe with the
same rules that work for most patients.

### Score interpretation

| Score | Interpretation | Suggested action |
|---|---|---|
| 0.0 - 0.4 | Typical — similar to cohort majority | Use as controls; calibrate your expectations |
| 0.4 - 0.6 | Moderately unusual — borderline | Review in cluster context |
| 0.6 - 1.0 | **Highly unusual** — top 5-10% | Prioritise for clinical review |

### What high anomaly means clinically

Possible causes of a high anomaly score:

- **Genuine clinical complexity** — multimorbid patient with atypical
  treatment history
- **Rare disease subtype** — presentation that differs from the dominant cohort
  pattern
- **Data coding difference** — institution-specific coding practice that
  differs from the majority
- **Data quality issue** — duplicate records, impossible date sequences,
  miscoded concepts
- **Short observation period** — patient recently enrolled, most windows empty

**None of these can be distinguished without chart review.** The anomaly score
is a prioritisation signal, not a diagnosis.

---

## 2. Temporal rule flags

### What they detect

Temporal flags are rule-based triggers applied to each patient's structured
OMOP record. They flag patterns that are worth reviewing — not patterns that
are definitively wrong.

### Current flags

```{r flags}
# Run temporal flags programmatically
flags <- detect_temporal_flags(
  cohort_members = cohort_members,
  domain_data    = domain_data,
  config         = temporal_flag_config(death_window_days = 90L)
)
```

| Flag type | Trigger | Most common benign explanation |
|---|---|---|
| `no_post_index_followup` | Zero events after index | Patient transferred care or data lag |
| `sparse_post_index` | Bottom 20% of cohort post-index events | Short follow-up or care elsewhere |
| `death_shortly_after_index` | Death within 90 days post-index | Valid outcome; check censoring rules |
| `drug_before_diagnosis_absent_after` | Drug pre-index, absent post | Treatment discontinuation |
| `recurrent_condition` | Same concept 3+ times within 30 days | Re-coding, chronic condition management |
| `isolated_index_code` | One condition code, no post-index support | Single administrative entry |
| `concentrated_activity` | >70% of events in one window | Discrete clinical episode |

### Language to use

- Say: *"Potential temporal inconsistency — flag for review"*
- Not: *"This is a data error"* or *"This patient should not be in the cohort"*

---

## 3. Cluster hypotheses

### How they are generated

The Hypothesis Panel compares feature prevalence between ML-defined cluster
pairs using:

- **Wilcoxon rank-sum test** for continuous features
- **Fisher's exact test** for binary features
- **BH correction** for multiple comparisons

### Critical limitation: circularity

Hypotheses are tested on the **same data used to define the clusters**. This
creates circularity — you will almost always find statistical differences
between clusters that were constructed to be different. These p-values are
**not** independent pre-specified statistical tests.

### How to use results

```{r hypotheses}
hypotheses <- generate_hypotheses(
  feature_matrix  = feature_matrix,
  ml_results      = ml_results,
  min_effect_size = 0.1,
  max_hypotheses  = 20L
)
# Sort by effect_size (largest difference) rather than p-value alone
hypotheses[order(-hypotheses$effect_size), ]
```

Use the hypothesis table to:

1. Identify concepts with large effect sizes (> 0.3) for clinical discussion
2. Ask a clinician: *"Does this difference between groups make clinical sense?"*
3. Design a formal pre-specified analysis in an independent dataset

### Required validation steps before reporting

1. Replicate in an independent cohort or time period
2. Use CohortDiagnostics to formally validate the phenotype
3. Pre-register the hypothesis before analysing the validation dataset

---

## 4. Summary: cautious language guide

| Output | What to say | What NOT to say |
|---|---|---|
| Anomaly score 0.85 | "Potential outlier — possible clinical complexity or data pattern" | "This patient is an error" |
| Cluster label "Medication-dense group" | "This group may reflect intensive treatment — requires review" | "This cluster is disease subtype X" |
| Temporal flag | "Potential temporal pattern of interest — review source records" | "This is a data quality failure" |
| Hypothesis | "Potential signal — hypothesis-generating, requires validation" | "We found that cluster A has more Y than B" |
