---
title: "CohortIntelligence — Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CohortIntelligence -- Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

CohortIntelligence is an OHDSI-native cohort exploration workbench for OMOP
CDM studies. This vignette walks through launching the dashboard in demo mode,
understanding the five-tab workflow, and running the core functions
programmatically.

## 1. Launch in demo mode

No database or credentials are required for demo mode:

```{r demo}
library(CohortIntelligence)
launch_cohort_intelligence()
```

The browser opens automatically with 50 synthetic patients loaded across all
seven OMOP domains. All five dashboard tabs are fully functional.

## 2. The five-tab workflow

| Tab | Question answered |
|---|---|
| Cohort Overview | What does my cohort look like at a glance? |
| Cluster & Anomaly | What subgroups exist, and who is unusual? |
| Review Queue | Which patients should I inspect first? |
| Trajectory Review | What is this patient's clinical history? |
| Hypothesis & Report | What findings can I bring to a clinician meeting? |

## 3. Run the pipeline programmatically

```{r pipeline}
library(CohortIntelligence)

# Use the built-in synthetic data
connector <- CohortIntelligence:::.cohort_intel_env  # demo only
# For real use, see the atlas-json-workflow vignette

# ── Step 1: create an in-memory connector from synthetic data
# (In production you would use create_cohort_connector())
```

## 4. Understand the output columns

```{r rank}
# After loading, rank_df has these key columns:
# subject_id    -- patient identifier
# rank_position -- 1 = highest priority for review
# priority_tier -- "high", "medium", "low"
# anomaly_score -- 0 (typical) to 1 (unusual); 0 if isotree not installed
# sparsity_score -- 0 (complete data) to 1 (all windows empty)
# cluster_id    -- ML cluster; -1/0 = unassigned (noise)
```

## 5. Signal explanation

```{r explain}
# Explain why a specific patient was prioritised
explanation <- explain_patient_priority(
  subject_id      = 1L,
  rank_df         = rank_df,
  feature_matrix  = feature_matrix,
  domain_activity = domain_activity,
  cohort_members  = cohort_members
)

# Each row is one explanation card
print(explanation[, c("explanation_label", "severity", "domain")])
```

## 6. Cautious language

All outputs are hypothesis-generating. Always use:

- "potential signal" (not "finding")
- "possible interpretation" (not "diagnosis")
- "requires clinical validation" (not "confirmed")
- "review priority" (not "clinical risk")

See `vignette("interpreting-results")` for a full guide.
